Chaotic Behavior of the Triple Pendulum \\ \large A Computational Approach

### 2017-05-09 paper.tex
further work

    \section{Equations of Motion}
    \subsection{Full Solution}

@@

        x_1 &= l\sin\phi_1     & y_1 &= l(1-\cos\phi_1) \\
        x_2 &= x_1+l\sin\phi_2 & y_2 &= y_1 + l(1-\cos\phi_2) \\
        x_3 &= x_2+l\sin\phi_3 & y_3 &= y_2 + l(1-\cos\phi_3).

@@

        y_1 &= l(1-\cos\phi_1) \\

@@

        y_2 &= l((1-\cos\phi_1)+(1-\cos\phi_2)) \\

@@

        y_3 &= l((1-\cos\phi_1)+(1-\cos\phi_2)+(1-\cos\phi_3)).

@@

    up $-mgy_i$ and $\frac{1}{2}mv_i^2$ for each mass $i$. We will drop the
    constant terms from the potential energy expression since they will be lost
    when we take the partial derivatives of the Lagrangian:

@@

        \L &= T + U \\
        &= -mgl\left(3\cos\phi_1+2\cos\phi_2+\cos\phi_3 \right) 

@@

        \pD{\L}{\phi_1} &= 3mgl\sin\phi_1 

@@

        \pD{\L}{\phi_2} &= 2mgl\sin\phi_2 

@@

        \pD{\L}{\phi_3} &= mgl\sin\phi_3 

@@

        3mgl\sin\phi_1 -2ml^2\dot\phi_1\dot\phi_2\sin(\phi_1-\phi_2)

@@

        3mgl\sin\phi_1 &= 3ml^2\ddot\phi_1 + 2ml^2\ddot\phi_2\cos(\phi_1-\phi_2)

@@

            - ml^2\dot\phi_3^2\sin(\phi_1-\phi_3)

@@

        \frac{3g}{l}\sin\phi_1 &= 3\ddot\phi_1 + 2\ddot\phi_2\cos(\phi_1-\phi_2)

@@

            - \dot\phi_3^2\sin(\phi_1-\phi_3)   \label{eq_m1}

@@

    For $\phi_2$ we have
    \begin{align}
        \pD{\L}{\phi_2} &= \D{}{t}\pD{\L}{\dot\phi_2} 
    \end{align}
    \begin{align}
        2mgl\sin\phi_2 +2ml^2\dot\phi_1\dot\phi_2\sin(\phi_1-\phi_2)
            -ml^2\dot\phi_2\dot\phi_3\sin(\phi_2-\phi_3)
        &= 2ml^2\ddot\phi_2 + 2ml^2\ddot\phi_1\cos(\phi_1-\phi_2) \\
        &\quad - 2ml^2\dot\phi_1\sin(\phi_1-\phi_2)(\dot\phi_1-\dot\phi_2) \\
        &\quad + ml^2\ddot\phi_3\cos(\phi_2-\phi_3) \\
        &\quad - ml^2\dot\phi_3\sin(\phi_2-\phi_3)(\dot\phi_2-\dot\phi_3) 
    \end{align}
    \begin{align}
        2mgl\sin\phi_2 &= 2ml^2\ddot\phi_2 + 2ml^2\ddot\phi_1\cos(\phi_1-\phi_2)
            + ml^2\ddot\phi_3\cos(\phi_2-\phi_3)  \\
            &\quad + 2ml^2\dot\phi_1^2\sin(\phi_1-\phi_2)
            - ml^2\dot\phi_3^2\sin(\phi_2-\phi_3)
    \end{align}
    \begin{align}
        \frac{2g}{l}\sin\phi_2 &= 2\ddot\phi_2 + 2\ddot\phi_1\cos(\phi_1-\phi_2)
            + \ddot\phi_3\cos(\phi_2-\phi_3) + 2\dot\phi_1^2\sin(\phi_1-\phi_2)
            - \dot\phi_3^2\sin(\phi_2-\phi_3)   \label{eq_m2}
    \end{align}

    For $\phi_3$ we have
    \begin{align}
        \pD{\L}{\phi_3} &= \D{}{t}\pD{\L}{\dot\phi_3} 
    \end{align}
    \begin{align}
        mgl\sin\phi_3 +ml^2\dot\phi_1\dot\phi_3\sin(\phi_1-\phi_3)
            +ml^2\dot\phi_2\dot\phi_3\sin(\phi_2-\phi_3)
        &= ml^2\ddot\phi_3 + ml^2\ddot\phi_1\cos(\phi_1-\phi_3) \\
        &\quad - ml^2\dot\phi_1\sin(\phi_1-\phi_3)(\dot\phi_1-\dot\phi_3) \\
        &\quad + ml^2\ddot\phi_2\cos(\phi_2-\phi_3) \\
        &\quad - ml^2\dot\phi_2\sin(\phi_2-\phi_3)(\dot\phi_2-\dot\phi_3) 
    \end{align}
    \begin{align}
        mgl\sin\phi_3 &= ml^2\ddot\phi_3 + ml^2\ddot\phi_1\cos(\phi_1-\phi_3)
            + ml^2\ddot\phi_2\cos(\phi_2-\phi_3)  \\
            &\quad + ml^2\dot\phi_1^2\sin(\phi_1-\phi_3)
            - ml^2\dot\phi_2^2\sin(\phi_2-\phi_3)
    \end{align}
    \begin{align}
        \frac{g}{l}\sin\phi_3 &= \ddot\phi_3 + \ddot\phi_1\cos(\phi_1-\phi_3)
            + \ddot\phi_2\cos(\phi_2-\phi_3) + \dot\phi_1^2\sin(\phi_1-\phi_3)
            - \dot\phi_2^2\sin(\phi_2-\phi_3)   \label{eq_m3}
    \end{align}

@@

             &= -2\dot\phi_2^2\sin(\phi_1-\phi_2) + \dot\phi_3^2\sin(\phi_1-\phi_3)
             +\frac{3g}{l}\sin\phi_1 \\

@@

            &= - 2\dot\phi_1^2\sin(\phi_1-\phi_2) + \dot\phi_3^2\sin(\phi_2-\phi_3)
            +\frac{2g}{l}\sin\phi_2 \\
        \ddot\phi_1\cos(\phi_1-\phi_3+ \ddot\phi_2\cos(\phi_2-\phi_3)) + \ddot\phi_3
            &= - \dot\phi_1^2\sin(\phi_1-\phi_3) + \dot\phi_2^2\sin(\phi_2-\phi_3)
            +\frac{g}{l}\sin\phi_3

@@

            \dot\phi_3^2\sin(\phi_1-\phi_3) -2\dot\phi_2^2\sin(\phi_1-\phi_2)
                +\frac{3g}{l}\sin\phi_1 \\
            \dot\phi_3^2\sin(\phi_2-\phi_3) - 2\dot\phi_1^2\sin(\phi_1-\phi_2)
                +\frac{2g}{l}\sin\phi_2 \\
            \dot\phi_2^2\sin(\phi_2-\phi_3) -\dot\phi_1^2\sin(\phi_1-\phi_3)
                +\frac{g}{l}\sin\phi_3

@@

    \subsection{Small-angle Approximation}

    To check the validity of our numerical solution, we would like an analytical
    solution. Clearly a full analytical solution is impossible. However, by
    assuming that $\phi_1,\phi_2$, and $\phi_3$ remain small, we can radically
    simplify the equations of motion to reach a closed-form expression for
    $\phi_1,\phi_2$, and $phi_3$ as a function of time.

    We begin by substituting $\cos\phi_i=1-\frac{1}{2}\phi_i^2$, the second
    order Maclaurin approximation for cosine, into our expression for potential
    energy. We also substitute $\dot\phi_i\dot\phi_j\cos(\phi_i-\phi_j)=\dot
    \phi_i\dot\phi_j$, since for the small product $\dot\phi_i\dot\phi_j$, the
    cosine term is negligible. This allows us to simplify our expression for
    kinetic energy as well. We find
    \begin{align}
        U &= -mgl\left(3(1-\tfrac{1}{2}\phi_1^2)+2(1-\tfrac{1}{2}\phi_2^2
            +(1-\tfrac{1}{2}\phi_3^2))\right) \\
        &\simeq mgl\left(\frac{3}{2}\phi_1^2+\frac{2}{2}\phi_2^2
            +\frac{1}{2}\phi_3^2\right) \tag*{(since the constant terms will be
            dropped when the derivative is taken)} \\\\
        T &= \frac{1}{2}ml^2(3\dot\phi_1^2+2\dot\phi_2^2+\dot\phi_1^2
            +4\dot\phi_1\phi_2+2\dot\phi_1\phi_3+2\dot\phi_2\phi_3),
    \end{align}
    so the Lagrangian is
    \begin{align}
        \L &= T+U \\
        &= \frac{1}{2}ml^2(3\dot\phi_1^2+2\dot\phi_2^2+\dot\phi_1^2
            +4\dot\phi_1\phi_2+2\dot\phi_1\phi_3+2\dot\phi_2\phi_3)
            -mgl\left(\frac{3}{2}\phi_1^2+\frac{2}{2}\phi_2^2
            +\frac{1}{2}\phi_3^2\right)
    \end{align}
    Taking derivatives as required, we have
    \begin{align}
        \pD{\L}{\phi_1} &= -3mgl\phi_1, & 
        \pD{\L}{\dot\phi_1} &= ml^2(3\dot\phi_1+4\dot\phi_2+2\dot\phi_3)
        &\Rightarrow\quad
        \D{}{t}\pD{\L}{\dot\phi_1} &= ml^2(3\ddot\phi_1+4\ddot\phi_2+2\ddot\phi_3) \\
                %
        \pD{\L}{\phi_2} &= -2mgl\phi_2, & 
        \pD{\L}{\dot\phi_2} &= ml^2(4\dot\phi_1+2\dot\phi_2+2\dot\phi_3)
        &\Rightarrow\quad
        \D{}{t}\pD{\L}{\dot\phi_3} &= ml^2(4\ddot\phi_1+2\ddot\phi_2+2\ddot\phi_3) \\
                %
        \pD{\L}{\phi_3} &= -mgl\phi_3, & 
        \pD{\L}{\dot\phi_3} &= ml^2(2\dot\phi_1+2\dot\phi_2+\dot\phi_3)
        &\Rightarrow\quad
        \D{}{t}\pD{\L}{\dot\phi_3} &= ml^2(2\ddot\phi_1+2\ddot\phi_2+\ddot\phi_3) \\
    \end{align}
    Applying our equilibrium condition $\D{}{t}\pD{\L}{\dot\phi_i}=\pD{\L}
    {\phi_i}$, we have our three equations of motion:
    \begin{align}
        ml^2(3\ddot\phi_1+4\ddot\phi_2+2\ddot\phi_3) &= -3mgl\phi_1 \\
        ml^2(4\ddot\phi_1+2\ddot\phi_2+2\ddot\phi_3) &= -2mgl\phi_2 \\
        ml^2(2\ddot\phi_1+2\ddot\phi_2+\ddot\phi_3) &= -mgl\phi_3
    \end{align}
    Cancelling the common term of $ml^2$, we have
    \begin{align}
        3\ddot\phi_1+4\ddot\phi_2+2\ddot\phi_3 &= -\frac{3g}{l}\phi_1 \\
        4\ddot\phi_1+2\ddot\phi_2+2\ddot\phi_3 &= -\frac{2g}{l}\phi_2 \\
        2\ddot\phi_1+2\ddot\phi_2+\ddot\phi_3  &= -\frac{g}{l}\phi_3
    \end{align}
    We can write this system of equations as a single matrix equation \[
        M\vec{\ddot\phi} = -G\vec{\phi} 
    \] where
    \begin{align}
        M &= \begin{pmatrix}3&4&2\\4&2&2\\2&2&1\end{pmatrix} &
        G &= \frac{g}{l}\begin{pmatrix}3&0&0\\0&2&0\\0&0&1\end{pmatrix}
    \end{align}
    We assume a periodic solution of the form \[
        \vec\phi = \Re \vec{z}, \quad\mathrm{where}\quad
        \vec{z} = \vec{a}\,\e^{\i\omega t}
    \] Taking the second derivative of $\vec{z}$, we find \[
        \vec{\ddot z} = -\omega^2\vec{a}\,\e^{\i\omega t} = -omega^2\vec{z}
    \] So our differential equation becomes \[
        -\omega^2 M\vec{a}\e^{\i\omega t} = -G\vec{a}\e^{\i\omega t},
    \] which by rearranging and cancelling the common exponential term yields \[
        (G-\omega^2M)\vec{a} = 0
    \] This equation will have nontrivial solutions when $\det(G-\omega^2M)=0$.
    Calculating the determinant (again with a computer algebra system), we find
    \begin{align}
        w^{6}+\frac{9g}{l}w^4+\frac{9g^2}{l^2}w^2-\frac{3g^3}{l^3} &= 0
    \end{align}

>>>>

    \section{Equations of Motion}

@@

        x_1 &= l\sin\phi_1     & y_1 &= -l\cos\phi_1 \\
        x_2 &= x_1+l\sin\phi_2 & y_2 &= y_1 + -l\cos\phi_2 \\
        x_3 &= x_2+l\sin\phi_3 & y_3 &= y_2 + -l\cos\phi_3.

@@

        y_1 &= -l\cos\phi_1 \\

@@

        y_2 &= -l(\cos\phi_1+\cos\phi_2) \\

@@

        y_3 &= -l(\cos\phi_1+\cos\phi_2+\cos\phi_3).

@@

    up $-mgy_i$ and $\frac{1}{2}mv_i^2$ for each mass $i$:

@@

        \L &= T - U \\
        &= mgl\left(3\cos\phi_1+2\cos\phi_2+\cos\phi_3 \right) 

@@

        \pD{\L}{\phi_1} &= -3mgl\sin\phi_1 

@@

        \pD{\L}{\phi_2} &= -2mgl\sin\phi_2 

@@

        \pD{\L}{\phi_3} &= -mgl\sin\phi_3 

@@

        -3mgl\sin\phi_1 -2ml^2\dot\phi_1\dot\phi_2\sin(\phi_1-\phi_2)

@@

        -3mgl\sin\phi_1 &= 3ml^2\ddot\phi_1 + 2ml^2\ddot\phi_2\cos(\phi_1-\phi_2)

@@

            + ml^2\dot\phi_3^2\sin(\phi_1-\phi_3)

@@

        -\frac{3g}{l}\sin\phi_1 &= 3\ddot\phi_1 + 2\ddot\phi_2\cos(\phi_1-\phi_2)

@@

            + \dot\phi_3^2\sin(\phi_1-\phi_3)   \label{eq_m1}

@@

    For $\phi_2$ we have
    \begin{align}
        \pD{\L}{\phi_2} &= \D{}{t}\pD{\L}{\dot\phi_2} 
    \end{align}
    \begin{align}
        -2mgl\sin\phi_2 +2ml^2\dot\phi_1\dot\phi_2\sin(\phi_1-\phi_2)
            -ml^2\dot\phi_2\dot\phi_3\sin(\phi_2-\phi_3)
        &= 2ml^2\ddot\phi_2 + 2ml^2\ddot\phi_1\cos(\phi_1-\phi_2) \\
        &\quad - 2ml^2\dot\phi_1\sin(\phi_1-\phi_2)(\dot\phi_1-\dot\phi_2) \\
        &\quad + ml^2\ddot\phi_3\cos(\phi_2-\phi_3) \\
        &\quad - ml^2\dot\phi_3\sin(\phi_2-\phi_3)(\dot\phi_2-\dot\phi_3) 
    \end{align}
    \begin{align}
        -2mgl\sin\phi_2 &= 2ml^2\ddot\phi_2 + 2ml^2\ddot\phi_1\cos(\phi_1-\phi_2)
            + ml^2\ddot\phi_3\cos(\phi_2-\phi_3)  \\
            &\quad - 2ml^2\dot\phi_1^2\sin(\phi_1-\phi_2)
            + ml^2\dot\phi_3^2\sin(\phi_2-\phi_3)
    \end{align}
    \begin{align}
        -\frac{2g}{l}\sin\phi_2 &= 2\ddot\phi_2 + 2\ddot\phi_1\cos(\phi_1-\phi_2)
            + \ddot\phi_3\cos(\phi_2-\phi_3) - 2\dot\phi_1^2\sin(\phi_1-\phi_2)
            + \dot\phi_3^2\sin(\phi_2-\phi_3)   \label{eq_m2}
    \end{align}

    For $\phi_3$ we have
    \begin{align}
        \pD{\L}{\phi_3} &= \D{}{t}\pD{\L}{\dot\phi_3} 
    \end{align}
    \begin{align}
        -mgl\sin\phi_3 +ml^2\dot\phi_1\dot\phi_3\sin(\phi_1-\phi_3)
            +ml^2\dot\phi_2\dot\phi_3\sin(\phi_2-\phi_3)
        &= ml^2\ddot\phi_3 + ml^2\ddot\phi_1\cos(\phi_1-\phi_3) \\
        &\quad - ml^2\dot\phi_1\sin(\phi_1-\phi_3)(\dot\phi_1-\dot\phi_3) \\
        &\quad + ml^2\ddot\phi_2\cos(\phi_2-\phi_3) \\
        &\quad - ml^2\dot\phi_2\sin(\phi_2-\phi_3)(\dot\phi_2-\dot\phi_3) 
    \end{align}
    \begin{align}
        -mgl\sin\phi_3 &= ml^2\ddot\phi_3 + ml^2\ddot\phi_1\cos(\phi_1-\phi_3)
            + ml^2\ddot\phi_2\cos(\phi_2-\phi_3)  \\
            &\quad - ml^2\dot\phi_1^2\sin(\phi_1-\phi_3)
            - ml^2\dot\phi_2^2\sin(\phi_2-\phi_3)
    \end{align}
    \begin{align}
        -\frac{g}{l}\sin\phi_3 &= \ddot\phi_3 + \ddot\phi_1\cos(\phi_1-\phi_3)
            + \ddot\phi_2\cos(\phi_2-\phi_3) - \dot\phi_1^2\sin(\phi_1-\phi_3)
            - \dot\phi_2^2\sin(\phi_2-\phi_3)   \label{eq_m3}
    \end{align}

@@

             &= -2\dot\phi_2^2\sin(\phi_1-\phi_2) - \dot\phi_3^2\sin(\phi_1-\phi_3)
             -\frac{3g}{l}\sin\phi_1 \\

@@

            &= 2\dot\phi_1^2\sin(\phi_1-\phi_2) - \dot\phi_3^2\sin(\phi_2-\phi_3)
            -\frac{2g}{l}\sin\phi_2 \\
        \ddot\phi_1\cos(\phi_1-\phi_3)+\ddot\phi_2\cos(\phi_2-\phi_3)+\ddot\phi_3
            &= + \dot\phi_1^2\sin(\phi_1-\phi_3) + \dot\phi_2^2\sin(\phi_2-\phi_3)
            -\frac{g}{l}\sin\phi_3

@@

            -\dot\phi_3^2\sin(\phi_1-\phi_3) -2\dot\phi_2^2\sin(\phi_1-\phi_2)
                -\frac{3g}{l}\sin\phi_1 \\
            -\dot\phi_3^2\sin(\phi_2-\phi_3) + 2\dot\phi_1^2\sin(\phi_1-\phi_2)
                -\frac{2g}{l}\sin\phi_2 \\
            \dot\phi_2^2\sin(\phi_2-\phi_3) + \dot\phi_1^2\sin(\phi_1-\phi_3)
                -\frac{g}{l}\sin\phi_3

@@
